Optimization

### 2020-11-28 optim/opt-talk/optimization.Rmd
WIP 28Nov2020

- minimize $f(x)$ subject to
  - $g_i(x) \ge 0$, $i = 1, \ldots, m$,
  - $h_i(x) = 0$, $i = 1, \ldots, n$.
- $f, g_i, h_i$ are arbitrary functions.
- $f$ is called the **objective function** while $g_i$ and $h_i$ are called
**constraints**.
- $x$ is an $n$-dimensional vector. That is, $x = (x_1, \ldots, x_n)$.
- This problem is usually quite difficult. 
- It has been attempted since at least $300$ years and has ushered many new concepts
in Mathematics.
- (Unfortunately) We are not mathematicians.
  - We will talk in the mathematical language but we will focus on the ideas, not
  the technicalities.
  - We will consider the optimization problem in simpler settings.

@@

- Suppose $x, y \in S$ then the set of points $\lambda x + (1 - \lambda)y$, where
$0 \le \lambda \le 1$ is a real number is called the set of all *convex combinations*
of $x$ and $y$.
  - In plain language, this set has all the points on the straight line joining $x$ and $y$.
- A set $S$ is called a convex set if convex combinations of all pairs of its points
lies in the set.
  - In plain language, if you take any two points in the set and draw a straight 
  line joining them then the line should lie entirely in the set.
- Examples of convex sets:
  - A circle, a square, an ellipse, a sphere, a box.
- Examples of sets that are not convex.
  - A circle with a hole, heart-shaped set, a torus (medu-vade)

@@

- A function $f$ is a rule that maps **every** member of a set $D$ to a **unique**
member of a set $R$. It is written as $f: D \mapsto R$. The set $D$ is called the
domain of the function and the set $R$ the range.
- Examples:
  - A rule that maps an integer to its square. Its domain is $\mathbb{Z}$ and its
  range is $\mathbb{N}$.
- Let $f:\mathbb{R}^m \mapsto \mathbb{R}^n$. $f$ is convex if 
$||f(\lambda x + (1 - \lambda y))|| \le \lambda ||f(x)|| + (1 - \lambda) ||f(y)||$.
- What is $||f(x)||$? $f(x) \in \mathbb{R}^n$. It is a vector. $||f(x)||$ is the
magnitude of the vector.
  - The magnitude of a vector $y = (y_1, \ldots, y_n)$ is $||y|| = \sqrt{y_1^2 + \cdots + y_n^2}$.
- If $f$ is convex, $-f$ is concave.
- A function may be convex in some region, concave in others.
- Choose $\lambda = 1/2$. If $\bar{x} = (x + y)/2$, $\overline{f(x)} = (f(x) + f(y))/2$ then
$||f(\bar{x})|| \le ||\overline{f(x)}||$.
  - Function of average is at most the average of the function.

@@

## Convex functions - 3
- Why are we so concerned about convex functions?
  - For a convex function, a local minimum is the global minimum.
  - If you find a minimum you can be assured that it is **the** minimum. You don't have to search any further.
- An optimization problem consists of
  - A set $F$ of feasible points.
  - A cost function $c: F \mapsto \mathbb{R}$.
- The solution of the optimization problem consists in finding $x^\ast \in F$ such that
$c(x^\ast) \le c(x)$ for all $x \in F$.
- A convex optimization problem is the one in which the set $F$ and the function $c$
are convex.
- In general, convex problems are much easier to solve than non-convex ones.
- CPLEX solves convex problems.

## Linear Programming
- Programming?
  - This is a term from the 1940s when a computer meant a person who computes and
  a program mean a plan.
  - In particular, it referred to the US army's training and logistic plans. George
  Dantzing, the man who developed the simplex algorithm was working on military
  problems during WW-II.
- It has the same meaning in 'dynamic programming', a style of building algorithms
that saves intermediate results in tables.
- In a linear programming problem the objective function and the constraints are
**linear functions**.
  - Constraints are in the form of lines/planes/hyperplanes.
  - Feasible region is a convex polygon/polyhedron/polytope.
- An optimum, if it exists, is at the corner of the polytope or along an edge.

>>>>

- minimize $f(x)$ subject to
  - $g_i(x) \ge 0$, $i = 1, \ldots, m$,
  - $h_i(x) = 0$, $i = 1, \ldots, n$.
- $f, g_i, h_i$ are arbitrary functions.
- $f$ is called the **objective function** while $g_i$ and $h_i$ are called
**constraints**.
- $x$ is an $n$-dimensional vector. That is, $x = (x_1, \ldots, x_n)$. These are the **decision variables**.
- This problem is usually quite difficult. 
- It has been attempted since at least $300$ years and has ushered many new concepts
in Mathematics.
- (Unfortunately) We are not mathematicians.
  - We will talk in the mathematical language but we will focus on the ideas, not
  the technicalities.
  - We will consider the optimization problem in simpler settings.

@@

- Suppose $x, y \in S$ then the set of points $\lambda x + (1 - \lambda)y$, where
$0 \le \lambda \le 1$ is a real number is called the set of all *convex combinations*
of $x$ and $y$.
  - In plain language, this set has all the points on the straight line joining $x$ and $y$.
- A set $S$ is called a convex set if convex combinations of all pairs of its points
lies in the set.
  - In plain language, if you take any two points in the set and draw a straight 
  line joining them then the line should lie entirely in the set.
- Examples of convex sets:
  - A circle, a square, an ellipse, a sphere, a box.
- Examples of sets that are [not convex](https://en.wikipedia.org/wiki/Convex_set).
  - A circle with a hole, heart-shaped set, a [torus](https://en.wikipedia.org/wiki/Torus).

@@

- A function $f$ is a rule that maps **every** member of a set $D$ to a **unique**
member of a set $R$. It is written as $f: D \mapsto R$. The set $D$ is called the
domain of the function and the set $R$ the co-domain.
- Examples:
  - A rule that maps an integer to its square. Its domain is $\mathbb{Z}$ and its
  codomain is $\mathbb{N}$.
- Let $f:\mathbb{R}^m \mapsto \mathbb{R}^n$. $f$ is convex if 
$||f(\lambda x + (1 - \lambda y))|| \le \lambda ||f(x)|| + (1 - \lambda) ||f(y)||$.
- What is $||f(x)||$? $f(x) \in \mathbb{R}^n$. It is a vector. $||f(x)||$ is the
magnitude of the vector.
  - The magnitude of a vector $y = (y_1, \ldots, y_n)$ is $||y|| = \sqrt{y_1^2 + \cdots + y_n^2}$.
- If $f$ is convex, $-f$ is concave.
- A function may be convex in some region, concave in others.
- Choose $\lambda = 1/2$. If $\bar{x} = (x + y)/2$, $\overline{f(x)} = (f(x) + f(y))/2$ then
$||f(\bar{x})|| \le ||\overline{f(x)}||$.
  - Function of average is at most the average of the function.

@@

## Convex functions - 3
- Why are we so concerned about convex functions?
  - For a convex function, a local minimum is the global minimum.
  - If you find a minimum you can be assured that it is **the** minimum. You don't have to search any further.
- An optimization problem consists of
  - A set $F$ of feasible points. A feasible point is the one that satisfies all constraints.
  - A cost function $c: F \mapsto \mathbb{R}$.
- The solution of the optimization problem consists in finding $x^\ast \in F$ such that
$c(x^\ast) \le c(x)$ for all $x \in F$.
- A convex optimization problem is the one in which the set $F$ and the function $c$
are convex.
- In general, convex problems are much easier to solve than non-convex ones.
- CPLEX solves convex problems.

## Linear Programming
- Programming?
  - This is a term from the 1940s when a computer meant a person who computes and
  a program mean a plan.
  - In particular, it referred to the US army's training and logistic plans. George
  Dantzing, the man who developed the simplex algorithm was working on military
  problems during WW-II.
- It has the same meaning in 'dynamic programming', a style of building algorithms
that save intermediate results in tables.
- In a linear programming problem the objective function and the constraints are
**linear functions**.
  - Constraints are in the form of lines/planes/hyperplanes.
  - Feasible region is a convex polygon/[polyhedron](https://en.wikipedia.org/wiki/Polyhedron)/polytope.
- An optimum, if it exists, is at the corner of the polytope or along an edge.
